Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-03 doc/paper/paper.tex

\newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}

@@

This is a paper about chaos.
% TODO: mention \texttt{echonn}

@@

In order to find the Lyapunov time, we only need to solve for the largest
LCE. To find this, we will use a method detailed in
\cite{viswanath1998lyapunov}.

% TODO: Lyapunov calculation
% TODO: Test with Circle yielding zero

@@

\subsection{Lorenz System}

% TODO: Image of Lorenz System

@@

Not all parameters for the Lorenz equation will produce chaotic behavior. In
Lorenz's article, he uses the parameter values $\sigma=10$, $\beta=8/3$, and
$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of
$0.90566$ \cite{viswanath1998lyapunov}. There are other parameters which
produce chaotic results outlined in Table \ref{table:lorenz_params}. Also
shown in Table \ref{table:lorenz_params} are the values calculated by my
Python module \texttt{echonn} are listed. These values are not as accurate,
but demonstrate the sufficient accuracy of \texttt{echonn}.

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|}
         % TODO: \lambda_1?
         % TODO: calculated values
        \hline
        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda$ & calculated $\lambda$ & relative error of calculated $\lambda$ \\
        \hline \hline
        16 & 45.92 & 4 & 1.50255 & & \\ % DATA POINTS
        16 & 40 & 4 & 1.37446 & & \\
        10 & 28 & 8/3 & 0.90566 & & \\
        \hline
    \end{tabular}
    \caption{
        Lorenz Parameters and Largest Lyapunov Exponent
        \cite{viswanath1998lyapunov}
    }
    \label{table:lorenz_params}
\end{table}

@@

Derived in \cite{eberle2007case}, the system of equations defined with
dimensionless rotating (synodic) coordinates:

@@

@@

Here, $x_1$ and $y_1$ represent the position of the third body relative to
the other two, but not in the standard Cartesian coordinates of 3D space.
Outlined in the paper \cite{eberle2007case}, the transformation back to
Cartesian coordinates is given, but we don't need the exact position of the
bodies since we are only concerned with the chaotic motion, so that isn't
shown here.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision.png}
        \caption{The restricted 3 body problem with collision}
        \label{fig:r3b_collision}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
        \caption{The resultant divergence of the body}
        \label{fig:r3b_collision_whole}
    \end{subfigure}
    \caption{The Restricted 3 Body Problem Error from Collision}
    \label{fig:singularity}
\end{figure}

@@

% TODO: position of the pendulum end points

>>>>

\newcommand{\echonn}{\texttt{echonn}}
\newcommand{\der}[2][t]{\frac{\mathrm{d}#2}{\mathrm{d}#1}}

@@

This is a paper about chaos.
% TODO: Intro
% TODO: mention \texttt{echonn}

@@

\subsection{Calculating the Largest Lyapunov Characteristic Exponent}

In order to find the Lyapunov time, we only need to solve for the largest
LCE. To find this, we will use the following method detailed in
\cite{viswanath1998lyapunov}.

\newcommand{\xv}{\mathbf{x}}
\newcommand{\fv}{\mathbf{f}}
\newcommand{\Mv}{\mathbf{M}}

The dynamical system will be defined by $\der{\xv}=\fv(\xv)$ where $\fv, \xv \in 
\mathbb{R}^d$. Given an initial condition $\xv_0$, the largest LCE, denoted 
$\lambda_1$, is calculated as follows:

\begin{equation}
    \lambda_1 = \lim_{t \to \infty} \frac{1}{t}\log \sigma_1(\frac{\partial \xv(t)}{\partial \xv_0}) \label{eq:max_lce}
\end{equation}

where $\sigma_1$ is the first singular value of the matrix $\frac{\partial 
\xv(t)}{\partial \xv_0}$ and $\log$ is the natural logarithm. Since we can only 
numerically approximate the limit for $\lambda_1$, we will need to use a
sufficiently large $t$ until the limit converges to the accuracy we need.

In order to calculate $\frac{\partial \xv(t)}{\partial \xv_0}$, we need to integrate the matrix differential equation:

\begin{equation}
    \der{\Mv(t)} = \frac{\partial \fv}{\partial \xv} \Mv(t)
    \label{eq:lce_mat_diffeq}
\end{equation}

where $\Mv(t) \in \mathbb{R}^{d \times d}$ and $\frac{\partial \fv}{\partial \xv}$
is the $d \times d$ Jacobian matrix of $\fv(\xv)$ at $\xv(t)$.

Since values for $\xv(t)$ are needed as well, the systems will be integrated 
together. If the dynamical system has dimensions $d$, then the new system of
differential equations will have dimension $d + d \times d$. For larger dynamical 
systems, this calculation becomes costly. The dynamical systems analyzed are
small enough (having dimensions ranging from 3-4) that the calculation can be 
performed timely and accurately.

There are several issues with calculating $\lambda_1$ with this algorithm.
Already mentioned, we need to choose a sufficiently
large $t$ for the limit to converge. 
Unfortunately, for chaotic systems, $\Mv$ will increase rapidly with time. The 
large floating point values will eventually overflow and $\lambda_1$ will not
be accurate. I have found empirically that if values of $\Mv$ don't exceed 
$10^{100}$, then $\lambda_1$ will be accurate.

Implemented in \echonn, the algorithm was tested on the Lorenz system for various 
parameters and initial conditions. The LCEs match the expected values very 
accurately (see the Lorenz System section: \ref{sec:lorenz}).
The 3 dimensional 3-body problem is an 18 dimensional dynamical system, so while 
the dynamical system is implemented in \echonn, its max LCE can't be calculated 
accurately since the resultant system has 342 dimensions. The algorithm was 
also tested on the differential equation which produces a circle:

\begin{align*}
    x' &= y \\
    y' &= -x
\end{align*}

with the solution

\begin{align*}
    x(t) &= r\sin(t) \\
    y(t) &= r\cos(t)
\end{align*}

for any $r \in \mathbb{R}$.

Given an initial condition $(x_0, y_0)$, it produces the trajectory of a circle
with radius $r=\sqrt{x_0^2 + y_0^2}$. If perturbed, the radius will change, 
$r_\epsilon=\sqrt{(x_0+\epsilon_x)^2 + (y_0+\epsilon_y)^2}$, so the new 
trajectory is

\begin{align*}
    x_\epsilon(t) &= r_\epsilon\sin(t) \\
    y_\epsilon(t) &= r_\epsilon\cos(t).
\end{align*}

Now, the $L^2$ norm of error is 
\begin{align*}
    \sqrt{(x_\epsilon - x)^2 + (y_\epsilon - y)^2} &= \sqrt{(r_\epsilon - r)^2\sin^2(t) + (r_\epsilon - r)^2\cos^2(t)} \\
    &= \sqrt{(r_\epsilon - r)^2(\sin^2(t) + \cos^2(t))^2} \\
    &= \sqrt{(r_\epsilon - r)^2} \\
    &= \vert r_\epsilon - r \vert.
\end{align*}

The error doesn't grow at all over time, so the LCE should be less than or equal to zero as the system is not chaotic. The algorithm reports an LCE of $-1.15954 \times 10^{-06}$ as expected.

Empirically shown, \echonn can be used to identify the max LCE of smaller 
dynamical systems, and special precaution has to be taken when choosing a 
time interval. Using a larger time will yield a more accurate $\lambda_1$, but 
it may also cause an overflow if the system is chaotic. Thus, 
maximum time values need to be chosen such that $\vert \vert \Mv\vert 
\vert_{\infty}$ doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert 
\vert_{\infty}$ is the $L^\infty$ norm).

@@

\subsection{Lorenz System}
\label{sec:lorenz}

@@

Not all parameters for the Lorenz equation will produce chaotic behavior. In
Lorenz's article, he uses the parameter values $\sigma=10$, $\beta=8/3$, and
$\rho=28$ \cite{lorenz1963deterministic}, and it has a maximum LCE of
$0.90566$ \cite{viswanath1998lyapunov}. There are other parameters which
produce chaotic results outlined in Table \ref{table:lorenz_params}. Also
shown in Table \ref{table:lorenz_params} are the values calculated by my
Python module \echonn are listed. These values are not as accurate,
but demonstrate the sufficient accuracy of \echonn.

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|}
         % TODO: \lambda_1?
         % TODO: calculated values
        \hline
        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda_1$ & calculated $\lambda_1$ & relative error of calculated $\lambda_1$ \\
        \hline \hline
        16 & 45.92 & 4 & 1.50255 & & \\ % DATA POINTS
        16 & 40 & 4 & 1.37446 & & \\
        10 & 28 & 8/3 & 0.90566 & & \\
        \hline
    \end{tabular}
    \caption{
        Lorenz Parameters and Largest Lyapunov Exponent
        \cite{viswanath1998lyapunov}
    }
    \label{table:lorenz_params}
\end{table}

The Lorenz system is named after its creator, but it is also known as the
Lorenz butterfly due to the shape of the attractor of the system. The Lorenz
Butterfly can be seen in Figure \ref{fig:lorenz_butter}.

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.6\textwidth}
        \includegraphics[width=\textwidth]{images/lorenz_nd_plot.png}
        \caption{Plot of each dimension of Lorenz System}
        \label{fig:lorenz_nd}
    \end{subfigure}
    \begin{subfigure}[b]{0.6\textwidth}
        \includegraphics[width=\textwidth]{images/lorenz_3d_plot.png}
        \caption{The Lorenz Butterfly}
        \label{fig:lorenz_3d}
    \end{subfigure}
    \caption{The Lorenz System Visuals}
    \label{fig:lorenz_butter}
\end{figure}

@@

Derived in two independent papers \cite{eberle2007case} and \cite{frnkacase},
the system of equations defined with dimensionless rotating (synodic)
coordinates:

@@

Although the notation in both papers was different, the equations were the same. 
It's also worth mentioning that Eberle's paper was for his Masters Thesis, and 
Frnka's paper has no details associated with it, but I was able to find a 
mathematician with the exact name who achieved his doctorate. With that said, 
I believe the math is reliable, and the simulations seem to work and at least 
demonstrate chaotic properties (the only requirement for this project). With my
lacking physics knowledge, I was unable to derive these equations myself from
the assumptions made by Henri Poincaré outlined in 
\cite{chenciner2000remarkable}.

@@

Here, $x_1$ and $y_1$ represent the position of the third body relative to
the other two, but not in the standard Cartesian coordinates of 3D space.
Outlined in the both papers, \cite{eberle2007case} and \cite{frnkacase}, the transformation back to
Cartesian coordinates is given, but we don't need the exact position of the
bodies since we are only concerned with the chaotic motion, so that isn't
shown here.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision.png}
        \caption{Restricted 3 Body Problem with collision}
        \label{fig:r3b_collision}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
        \caption{The resultant divergence of the body}
        \label{fig:r3b_collision_whole}
    \end{subfigure}
    \caption{The Restricted 3 Body Problem Error from Collision}
    \label{fig:singularity}
\end{figure}

@@
